Supplementary Materials- A Bayesian networks approach to infer social changes in burials in northeastern Taiwan during the European colonization period

### 2021-06-05 analysis/paper/supplementary-material.Rmd
edit the bib file and the SI rmd
edit the SI

image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
image_width_2_col_inch = 190/25.4 # image Double column w = 190 mm 
```

@@

<!--fit in a bigger historical picture in a pericolonial context: move to SI -->
The involvement in long-distance exchange is often associated with changes in political-economic strategies from a corporate mode to a network mode, where ambitious individuals are able to build their local power base and personal influence through the acquisition and distribution of foreign prestige goods [@Blanton1996; @Feinman2000; @Klehm2017]. Increased social inequality associated with the use of exotic trade goods is recognized as a ‘network’ strategy for gaining power [@Blanton1996; @Feinman2000]. Northeastern Taiwan became more involved in regional exchange with East Asia since the 14th century and long-distance trade with Southeast Asia brought by Europeans in the 17th century [@Chen2005; @Wang2007]. In 1626, the Spanish built their forts in Heping dao and Tamsui, northern Taiwan, and later were defeated and taken over by the Dutch in 1642 [@Andrade2007]. Indigenous people in northeastern Taiwan first encountered colonial power in 1632 when they were attacked by the Spanish and later in 1647 by the Dutch. The Dutch asked local settlements to pay deer hides or rice as annual tributes but negotiating a delayed payment is allowed if it is needed [@Borao2009; @Kang2016]. In the annual event held by the Dutch with participants of local Indigenous representatives [@Andrade1997], the Dutch would give those representatives symbolic items such as rattan sticks or foreign goods to assert their political authority and consolidate their relationship with Indigenous settlements. The introduction of foreign prestige goods to local Indigenous communities might have occurred before European presence became established, but was maintained and amplified after European arrival due to local competition for power [@Borao2009; @LiandWu2006; @LWandBM2020ornament; @Ueda2016].

```{r import-and-inspect-data, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/000-prep-data.R"))}
suppressWarnings(source(here::here("analysis", "paper", "000-prep-data.R")))
```

```{r tidy-data, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/001-data-tidy.R"))}
suppressWarnings(source(here::here("analysis", "paper", "001-data-tidy.R")))
```

```{r burial-location, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/002-burial-location.R"))}
suppressWarnings(source(here::here("analysis", "paper", "002-burial-location.R")))
```

@@

<!-- intro to networks: move to SI -->
Drawing largely on network science and graph theory, a key assumption in the construction of archaeological networks is that a phenomenon can be conceptualized as a network through abstraction and representation [@Brandes2013; @Collar2015; @Mills2017; @Peeples2019]. A social network is generally visualized as a graph consisting of a set of socially-relevant nodes/actors, connected by edges/ties representing one or more relations, such as friendship, collaborations, information flow, trade ties, or any other forms of connection of interest [@Marin2011; @Wasserman1994]. The ties can be classified into four major types, including similarities, social relations, interactions, and flows [@Borgatti2009]. In archaeology, actors can be people, groups, objects, places, or events, with ties built on similarity, proximity, or co-presence of material culture to create patterns reflecting influence, geographical distance, or affiliations in social groups [@Mills2017; @Brughmans2018; @Peeples2019]. For example, past trade can be conceptualized as a network of individual entities connected by shared similar objects or elements, such as the flow of goods, to represent their interactions [@Collar2015, pp. 4]. This concept can be applied to burial contexts where each burial is an actor linked by similar elements, such as burial goods. Burial goods, especially high value goods, can reflect social practices in broader cultural contexts to represent personal wealth or social status from which we can infer social differentiation or complexity [@Gamble2002; @Janes2013]. This enables the exploration of the structure of the past social organization through the identification of the relationships among burials.

<!-- two types of network analysis: move to SI -->
Network analysis has been increasingly applied by archaeologists in recent years to deal with past interactions and explore the underlying mechanisms. There are two common approaches to characterize network properties at two distinct scales: node/edge level and graph level [@Peeples2019]. Node level focuses on node properties in a network, such as centrality, representing the individual influence or social prominence in a group, while graph level assesses the whole network attributes, such as density, clustering in a network, to generalize relationship patterns [@Mills2017; @Peeples2019]. By quantifying those network properties, archaeologists can answer a wide range of research questions.

@@

<!-- current approaches to network analysis: move to SI -->
Current approaches to network analysis used by archaeologists are mostly restricted to a single rational structure without consideration of interaction between network variables. Our use of Bayesian inference on ERGMs is the first application to archaeological data that can bring new insights to understand past social structures by characterizing network properties as a whole. ERGMs are an important family of statistical models for networks that allows direct modeling for the formation of edges, or ties, between nodes [@Robins2007]. The assumption is that possible ties in a network are random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. In other words, networks in ERGMs are viewed as dependent variables, where network dependencies and the attributes of nodes/edges can influence the formation of a tie [@Snijders2011]. For example, nodes with similar attributes are more likely to have a relationship, such as friendship between people with the same hobby. Ties form a small structure in a network called a graph configuration, that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity or clustering (relationship between two actors through a shared third actor), homophily (relationship between actors with a similar attribute), and popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations represent the structure or the property of a network and can be expressed by network statistics.

@@

<!-- burial data and the definition of tie formation: partially move to SI  -->
For trade beads with substantial differences in quantities between burials, we described each burial as having one of four levels according to their distributions across all burials (Figure \@ref(fig:figure-raincloud-beads), also in the main text). Gold-foil beads are in levels of high (\>3), upper-middle (3), lower-middle (2), and low (1); carnelian beads are in levels of high (\>6), upper-middle (4-6), lower-middle (3), and low (1-2); glass beads are in levels of high (\>6), upper-middle (3-6), lower-middle (2), and low (1). If burial 1 and burial 2 both have high quantities of agate beads, then there will be a tie connecting them. The trade goods we used to define network ties are relatively durable materials, such as glass, stone, and metal, that are less affected by taphonomic factors in buried environments. In addition, each burial has a clear boundary as our analytical unit that reduces sampling errors. Thus, we are confident that the quantities and distributions of beads in burial contexts can reflect the original differences between burials. Despite the exclusion of eight burials lacking clear archaeological contexts, the burials analyzed in this study consisting of 90% Kiwulan burials can effectively represent structures of burial networks.

For burial attributes, we assigned burials to different groups according to their age, sex, burial value and the presence/absence of ritual pottery. We have three groups for the age attribute, 0-12, 12-20, and above 20; two groups for the sex attribute, male and female; and four groups for the burial value that are specified in the main paper document. Table \@ref(tab:age-sex-table) shows the number of burials, of which sex and age can be determined.

```{r raincloud-beads, fig.cap = "The distributions of different types of trade beads across all burials analyzed in this paper."}
# combine ridge plots 1 & 4 in the 000-prep-data file
library(tidyverse)
library(ggridges)
ridge_1_and_4 <-
  burial_beads %>%
  mutate(type = case_when(
    type == "Golden_bead" ~"gold-foil bead",
    type == "all_glass_bead" ~"glass bead",
    type == "Agate_bead" ~"carnelian bead")) %>% 
  ggplot(aes(x = value,
             y = type,
             fill = factor(stat(quantile)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE,
    jittered_points = TRUE,
    alpha = 0.7,
    vline_size = 0.5,
    vline_color = "grey10",
    point_size = 2.5,
    point_alpha = 0.4,
    position = position_raincloud(adjust_vlines = TRUE),
    rel_min_height = 0.01,
    scale = 0.5, # so the filled regions don't overlap on the points
    size = 0.2) +
  scale_x_continuous(limits = c(1, 15),
                     expand = c(0.1, 0)) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_minimal()

#ggsave(here::here("analysis", "figures", "000-raincloud-beads.png"),
       #w = 6, h = 4)
```

@@

```{r figure-raincloud-beads, fig.cap = "The distributions of different types of trade beads across burials analyzed in this paper."}

knitr::include_graphics(here::here("analysis", "figures", "000-raincloud-beads.png"))
```

@@

```{r bergm-iteration}
burn_in <- 100
main_iter <- 1000
aux_iter <- 4000
bgof_si <-10000
```

@@

```{r bergm-pre, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/003-burials-pre-network-modelling.R"))}
suppressWarnings(source(here::here("analysis", "paper", "003-burials-pre-network-modelling.R")))
```

```{r bergm-post, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/004-burials-post-network-modelling.R"))}
suppressWarnings(source(here::here("analysis", "paper", "004-burials-post-network-modelling.R")))
```

>>>>

```

@@

<!--fit in a bigger historical picture in a pericolonial context: move to SI-->
The involvement in long-distance exchange is often associated with changes in political-economic strategies from a corporate mode to a network mode, where ambitious individuals are able to build their local power base and personal influence through the acquisition and distribution of foreign prestige goods [@Blanton1996; @Feinman2000; @Klehm2017]. Increased social inequality associated with the use of exotic trade goods is recognized as a ‘network’ strategy for gaining power [@Blanton1996; @Feinman2000]. Northeastern Taiwan became more involved in regional exchange with East Asia since the 14th century and long-distance trade with Southeast Asia brought by Europeans in the 17th century [@Chen2005; @Wang2007]. In 1626, the Spanish built their forts in Heping dao and Tamsui, northern Taiwan, and later were defeated and taken over by the Dutch in 1642 [@Andrade2007]. Indigenous people in northeastern Taiwan first encountered colonial power in 1632 when they were attacked by the Spanish and later in 1647 by the Dutch. The Dutch asked local settlements to pay deer hides or rice as annual tributes but negotiating a delayed payment is allowed if it is needed [@Borao2009; @Kang2016]. In the annual event held by the Dutch with participants of local Indigenous representatives [@Andrade1997], the Dutch would give those representatives symbolic items such as rattan sticks or foreign goods to assert their political authority and consolidate their relationship with Indigenous settlements. The introduction of foreign prestige goods to local Indigenous communities might have occurred before European presence became established, but was maintained and amplified after European arrival due to local competition for power [@Borao2009; @LiandWu2006; @LWandBM2020ornament; @Ueda2016].

```{r read-models-data, include=FALSE}
load(here::here("analysis", "data", "derived_data", "burial_bergm_model.RData"))
readRDS(here::here("analysis", "data", "derived_data", "pre_bergm.rds"))
readRDS(here::here("analysis", "data", "derived_data", "post_bergm.rds"))
```

@@

<!--intro to networks: moved from the main text-->
Drawing largely on network science and graph theory, a key assumption in the construction of archaeological networks is that a phenomenon can be conceptualized as a network through abstraction and representation [@Brandes2013; @Collar2015; @Mills2017; @Peeples2019]. A social network is generally visualized as a graph consisting of a set of socially-relevant nodes/actors, connected by edges/ties representing one or more relations, such as friendship, collaborations, information flow, trade ties, or any other forms of connection of interest [@Marin2011; @Wasserman1994]. The ties can be classified into four major types, including similarities, social relations, interactions, and flows [@Borgatti2009]. In archaeology, actors can be people, groups, objects, places, or events, with ties built on similarity, proximity, or co-presence of material culture to create patterns reflecting influence, geographical distance, or affiliations in social groups [@Mills2017; @Brughmans2018; @Peeples2019]. For example, past trade can be conceptualized as a network of individual entities connected by shared similar objects or elements, such as the flow of goods, to represent their interactions [@Collar2015, pp. 4]. This concept can be applied to burial contexts where each burial is an actor linked by similar elements, such as burial goods. Burial goods, especially high value goods, can reflect social practices in broader cultural contexts to represent personal wealth or social status from which we can infer social differentiation or complexity [@Gamble2002; @Janes2013]. This enables the exploration of the structure of the past social organization through the identification of the relationships among burials.

<!--two types of network analysis: moved from the main text-->
Network analysis has been increasingly applied by archaeologists in recent years to deal with past interactions and explore the underlying mechanisms. There are two common approaches to characterize network properties at two distinct scales: node/edge level and graph level [@Peeples2019]. Node level focuses on node properties in a network, such as centrality, representing the individual influence or social prominence in a group, while graph level assesses the whole network attributes, such as density, clustering in a network, to generalize relationship patterns [@Mills2017; @Peeples2019]. By quantifying those network properties, archaeologists can answer a wide range of research questions.

@@

<!--current approaches to network analysis: move from the main text-->
Current approaches to network analysis used by archaeologists are mostly restricted to a single rational structure without consideration of interaction between network variables. Our use of Bayesian inference on ERGMs is the first application to archaeological data that can bring new insights to understand past social structures by characterizing network properties as a whole. ERGMs are an important family of statistical models for networks that allows direct modeling for the formation of edges, or ties, between nodes [@Robins2007]. The assumption is that possible ties in a network are random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. In other words, networks in ERGMs are viewed as dependent variables, where network dependencies and the attributes of nodes/edges can influence the formation of a tie [@Snijders2011]. For example, nodes with similar attributes are more likely to have a relationship, such as friendship between people with the same hobby. Ties form a small structure in a network called a graph configuration, that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity or clustering (relationship between two actors through a shared third actor), homophily (relationship between actors with a similar attribute), and popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations represent the structure or the property of a network and can be expressed by network statistics.

@@

<!-- burial data and the definition of tie formation: partially move to SI  -->
For trade beads with substantial differences in quantities between burials, we described each burial as having one of four levels according to their distributions across all burials (Figure \@ref(fig:figure-raincloud-beads), also in the main text). Gold-foil beads are in levels of high (\>3), upper-middle (3), lower-middle (2), and low (1); carnelian beads are in levels of high (\>6), upper-middle (4-6), lower-middle (3), and low (1-2); glass beads are in levels of high (\>6), upper-middle (3-6), lower-middle (2), and low (1). If burial 1 and burial 2 both have high quantities of carnelian beads, then there will be a tie connecting them. The trade goods we used to define network ties are relatively durable materials, such as glass, stone, and metal, that are less affected by taphonomic factors in buried environments. In addition, each burial has a clear boundary as our analytical unit that reduces sampling errors. Thus, we are confident that the quantities and distributions of beads in burial contexts can reflect the original differences between burials.

For burial attributes, we assigned burials to different groups according to their age, sex, burial value and the presence/absence of ritual pottery. We have three groups for the age attribute, 0-12, 12-20, and above 20; two groups for the sex attribute, male and female; and four groups for the burial value that are specified in the main paper document. Table \@ref(tab:age-sex-table) shows the number of burials, of which sex and age can be determined. Despite the exclusion of eight burials lacking clear archaeological contexts, the burials analyzed in this study consisting of 90% Kiwulan burials can effectively represent structures of burial networks.

@@

@@

```{r bergm-iteration}
burn_in <- 1000
main_iter <- 40000
aux_iter <- 5000
bgof_si <-10000
```

@@
